Attribution of undesirable character traits, rather than trait-based dehumanization, predicts punishment decisions

### 2024-07-06 manuscript.Rmd
Update manuscript.Rmd

@Bastian2013 conducted influential empirical studies testing the hypothesised association between the denial of human character traits and the endorsement of harsh punishment (see also @Barber2022; @Chen-Xia2023; @Kasper2022; @Morehouse2023; @Rousseau2023; @West2022). The researchers measured how trait-based dehumanization influenced participants' punishment of criminals. Participants were asked to rate their agreement with four items assessing animalistic dehumanization of criminals: "*I felt like the person in the story was refined and cultured*" [reversed], "*I felt like the person in the story was rational and logical, like they were intelligent*" [reversed] "*I felt like the person in the story lacked self-restraint, like an animal*", and "*I felt like the person in the story was unsophisticated*". Participants were also asked to rate their agreement with four items assessing mechanistic dehumanization of criminals: "*I felt like the person in the story was open minded, like they could think clearly about things*" [reversed], "*I felt like the person in the story was emotional, like they were responsive and warm*" [reversed], "*I felt like the person in the story was superficial like they had no depth*", "*I felt like the person in the story was mechanical and cold, like a robot*". Bastian and colleagues reported that both forms of dehumanization predicted endorsement of harsh punishment for the criminals portrayed in their stimuli, concluding that their participants viewed criminals as '*subhuman and beastly*' (pp.9).

@@

@Bastian2013 seek to account for this possibility by statistically controlling for participants’ moral outrage at the targets’ behaviour in their analysis.  They report that the relationship between trait-based dehumanization and punishment remains even when moral outrage is controlled for.  While this is interesting and suggestive of the independent effects of dehumanization, it cannot fully address the conceptual weaknesses in how dehumanization was operationalised.  A more convincing way to de-confound evidence for trait-based dehumanization from evidence of negative evaluation is to ask participants to rate the target group on traits that are uniquely human but vary from socially desirable to undesirable [@Over2021a, @Over2021b].  Previous research conducted by @Enock2021b has established that undesirable character traits such as jealous, spiteful and bitter are considered unique to humans and socially undesirable.  Across three intergroup contexts, the researchers found that participants attributed socially desirable human traits more strongly to the ingroup and socially undesirable traits more strongly to the outgroup (see also @Decker2023; @Enock2021a; @Enock2022; @Enock2023).  @Enock2021b concluded that intergroup preference may better explain apparent evidence for trait-based dehumanization.  However, it is not yet clear how the attribution of uniquely human character traits relates to harm.  Addressing this question is crucial to understanding the extent to which the dual model of dehumanization can help explain real-world discrimination and negativity.

We revisit the hypothesised causal relationship between trait-based dehumanization and harm in the context of endorsing harsh punishment for criminals. In Studies 1A and 1B, we seek to conceptually replicate the key findings of Bastian and colleagues [-@Bastian2013], suggesting that the extent to which participants animalistically (Study 1A) and mechanistically (Study 1B) dehumanize criminals predicts the severity of the punishment participants endorse for them. In Studies 2A and 2B, we adopt a similar design but incorporate socially undesirable traits into our stimulus set. This addition to the design allows us to investigate whether trait-based dehumanization, undesirable trait attribution, or both predict the severity of punishment. Following Bastian and colleagues, and to understand the generalisability of our findings, we investigate these questions in relation to two different types of crime (violent crime and theft). In Studies 3A and 3B, we seek to investigate a similar question using an experimental design. We present participants with vignettes in which criminals are described using character traits that differ in how socially desirable they are and whether or not they are unique to humans. We also measure how these varying descriptions influence participants' parole decisions. This allows us to directly measure whether there is a causal relationship between trait-based dehumanization and punishment, independent of ingroup preference.

@@

@Bastian2013 presented evidence that the more participants dehumanize violent criminals, the harsher the punishment participants endorse for them. We sought to test whether we could conceptually replicate this relationship between trait-based dehumanization and punishment using terms very similar to those used by Bastian and colleagues. In study 1A, participants read a series of scenarios describing fictitious criminals and their violent crimes. Following this, participants rated the extent to which the criminals possessed four character traits that distinguish humans from animals (*refined, rational and logical, has a sense of morality* and *civilised*). Following common practice in the field, we refer to these as uniquely human traits. Participants also rated the extent to which criminals possessed four character traits that distinguish humans from machines (*openminded, emotionally responsive, has a depth of character*, and *interpersonally warm*). Following common practice in the field, we refer to these as human nature traits. Participants also responded to an item measuring how harsh they thought the violent criminals' punishment should be. Following @Bastian2013, we predicted that the less participants attributed these uniquely human traits to criminals, the harsher the punishment they would recommend for criminals.

@@

**Trait attribution**. After reading each vignette, participants responded to items designed to measure trait-based dehumanization, broadly following the procedure of Bastian and colleagues [-@Bastian2013]. Participants indicated the extent to which they attributed four uniquely human traits (*refined, rational and logical, has a sense of morality* and *civilised*) and four human nature traits (*openminded, emotionally responsive, has a depth of character*, and *interpersonally warm*) to the criminals depicted. Participants indicated their agreement with each item (e.g., '*I think* [e.g., *Charlie*] *is refined*') using an unmarked sliding scale from 0 ('*Strongly Disagree*') to 100 ('*Strongly Agree*'), with the sliders initially fixed at the midpoint. According to the dual model, lower scores indicated greater dehumanization of violent criminals. An attention check appeared halfway through the dehumanization items for two criminals ('*Please move the slider all the way to Strongly Agree/Disagree*').

@@

Following @Bastian2013, we utilised a within-subjects, correlational design. All participants read the same five vignettes presented in a random order and responded to the same trait attribution and punishment items. Participants' scores for the trait attribution items and the endorsed harshness of punishment item were then averaged across scenarios. The presentation of the items in the trait attribution task was also randomised.

@@

In line with our pre-registered criteria, seven highly influential were omitted from the analysis (remaining sample *N* = 93). After calculating the average attribution score for human nature traits and punishment for each participant, we conducted a simple linear regression to test attribution of human nature traits predicted the harshness of punishment endorsed for violent criminals. A significant negative relationship was found, *b* = `r round(Exp1a2$coefficients[2], digits = 2)` [`r round(Exp1a2CIdf$X2.5[2], digits = 2)`, `r round(Exp1a2CIdf$X97.5[2], digits = 2)`], *t* = `r round(Exp1a2$coefficients[6], digits = 2)`, *p* `r format.pval(Exp1a2$coefficients[8], digits = 2, eps = 0.001, nsmall = 3)`, see Figure \@ref(fig:figure1). This relationship shows that greater mechanistic dehumanization (operationalised as the denial of human nature traits) was associated with the endorsement of harsher punishment. The model explains `r round(Exp1a2$r.squared, digits = 2)*100`% of the variance in the harshness of punishment scores, $R^2$ = `r round(Exp1a2$r.squared, digits = 2)`, *F*(`r round(Exp1a2$fstatistic[2], digits = 2)`, `r round(Exp1a2$fstatistic[3], digits = 2)`) = `r round(Exp1a2$fstatistic[1], digits = 2)`.

@@

Study 1B investigates whether the relationship found in Study 1A replicates when participants are asked to judge a different type of criminal activity. In Study 1B, we examined whether animalistic and mechanistic dehumanization, as operationalised by @Bastian2013, are associated with the harshness of punishment endorsed for individuals who commit theft. The design, materials and analysis plan were similar to that used in Study 1A, except that the scenarios involved theft rather than violent crime.

@@

**Vignettes** All participants responded to the same five vignettes, each detailing a crime involving theft (see Supplementary materials). As in Study 1A, an effort was made to ensure the vignettes were similar in structure and amount of detail. Once again, all of the perpetrators had gender-neutral names. An example of one of the theft vignettes is as follows: "*Until their recent arrest, Charlie had worked as a till operator at a local charity shop supporting individuals experiencing homelessness. Charlie had been stealing cash amounts varying from £5 to £50 from the tills almost daily over a five-year period. Police revealed that Charlie had stolen several thousand pounds from the charity shop while working there*."

@@

#looks like there some outliers (over 3 standardised residuals)
#several scores have high leverage 2(3+1)/130 = .06.

@@

#plot(HN_harm, 5) #looks like there some outliers (over 3 standardised residuals)
#several scores have high leverage 2(3+1)/130 = .06.

@@

#Rename The columns in both data frames so they match and allow for merging
names(Study2A_processed) <-c("CaseID", "Desirability_condition", "Consent", "English_Fluency", "Over_18", 
                             "Age", "Gender", "Gender_text", "Been_in_prison", "Criminal1_UH1_cultured.corrupt", 
                             "Criminal1_UH2_civilised.controlling", "Criminal1_UH3_sophisticated.arrogant", 
                             "Criminal1_UH4_moral.bitter", "Criminal1_HN1_generous.jealous", "Criminal1_HN2_open-minded.selfish", 
                             "Criminal1_HN3_warm.spiteful", "Criminal1_HN4_kind.cruel", "Criminal1_harm", 
                             "Criminal2_UH1_cultured.corrupt", "Criminal2_UH2_civilised.controlling", 
                             "Criminal2_UH3_sophisticated.arrogant", "Criminal2_UH4_moral.bitter", 
                             "Criminal2_HN1_generous.jealous", "Criminal2_HN2_open-minded.selfish", "Criminal2_HN3_warm.spiteful", 
                             "Criminal2_HN4_kind.cruel", "Criminal2_harm", "Criminal3_UH1_cultured.corrupt", 
                             "Criminal3_UH2_civilised.controlling", "Criminal3_UH3_sophisticated.arrogant", 
                             "Criminal3_UH4_moral.bitter", "Criminal3_HN1_generous.jealous", "Criminal3_HN2_open-minded.selfish", 
                             "Criminal3_HN3_warm.spiteful", "Criminal3_HN4_kind.cruel", "Criminal3_harm", "Criminal4_UH1_cultured.corrupt", 
                             "Criminal4_UH2_civilised.controlling", "Criminal4_UH3_sophisticated.arrogant", "Criminal4_UH4_moral.bitter", 
                             "Criminal4_HN1_generous.jealous", "Criminal4_HN2_open-minded.selfish", "Criminal4_HN3_warm.spiteful", 
                             "Criminal4_HN4_kind.cruel", "Criminal4_harm", "Criminal5_UH1_cultured.corrupt", 
                             "Criminal5_UH2_civilised.controlling", "Criminal5_UH3_sophisticated.arrogant", 
                             "Criminal5_UH4_moral.bitter", "Criminal5_HN1_generous.jealous", "Criminal5_HN2_open-minded.selfish", 
                             "Criminal5_HN3_warm.spiteful", "Criminal5_HN4_kind.cruel", "Criminal5_harm", "UH_mean", "HN_mean", 
                             "Harm_mean", "Outlier_Animalistic_model", "Outlier_Mechanistic_model", "Study")

names(Study2B_processed) <-c("CaseID", "Desirability_condition", "Consent", "English_Fluency", "Over_18", 
                              "Age", "Gender", "Gender_text", "Been_in_prison", "Criminal1_UH1_cultured.corrupt", 
                              "Criminal1_UH2_civilised.controlling", "Criminal1_UH3_sophisticated.arrogant", 
                              "Criminal1_UH4_moral.bitter", "Criminal1_HN1_generous.jealous", "Criminal1_HN2_open-minded.selfish", 
                              "Criminal1_HN3_warm.spiteful", "Criminal1_HN4_kind.cruel", "Criminal1_harm", 
                              "Criminal2_UH1_cultured.corrupt", "Criminal2_UH2_civilised.controlling", 
                              "Criminal2_UH3_sophisticated.arrogant", "Criminal2_UH4_moral.bitter", 
                              "Criminal2_HN1_generous.jealous", "Criminal2_HN2_open-minded.selfish", "Criminal2_HN3_warm.spiteful", 
                              "Criminal2_HN4_kind.cruel", "Criminal2_harm", "Criminal3_UH1_cultured.corrupt", 
                              "Criminal3_UH2_civilised.controlling", "Criminal3_UH3_sophisticated.arrogant", 
                              "Criminal3_UH4_moral.bitter", "Criminal3_HN1_generous.jealous", "Criminal3_HN2_open-minded.selfish", 
                              "Criminal3_HN3_warm.spiteful", "Criminal3_HN4_kind.cruel", "Criminal3_harm", "Criminal4_UH1_cultured.corrupt", 
                              "Criminal4_UH2_civilised.controlling", "Criminal4_UH3_sophisticated.arrogant", "Criminal4_UH4_moral.bitter", 
                              "Criminal4_HN1_generous.jealous", "Criminal4_HN2_open-minded.selfish", "Criminal4_HN3_warm.spiteful", 
                              "Criminal4_HN4_kind.cruel", "Criminal4_harm", "Criminal5_UH1_cultured.corrupt", 
                              "Criminal5_UH2_civilised.controlling", "Criminal5_UH3_sophisticated.arrogant", 
                              "Criminal5_UH4_moral.bitter", "Criminal5_HN1_generous.jealous", "Criminal5_HN2_open-minded.selfish", 
                              "Criminal5_HN3_warm.spiteful", "Criminal5_HN4_kind.cruel", "Criminal5_harm", "UH_mean", "HN_mean", 
                              "Harm_mean", "Outlier_Animalistic_model", "Outlier_Mechanistic_model", "Study")

@@

Study 2A investigated whether apparent evidence for a relationship between trait-based dehumanization and endorsement of harsh punishment for violent criminals remains when controlling for the desirability of the traits. We tested this by introducing character traits perceived as uniquely human yet socially undesirable into the stimulus set [@Enock2021a; @Over2021a]. The dual model predicts that to the extent criminals are denied uniquely human character traits, they will be subjected to harsher punishment. We predict that trait desirability will moderate the relationship between human trait attribution and punishment. More specifically, we predict that the extent to which violent criminals are denied socially desirable character traits, and attributed socially undesirable character traits, will predict harsh punishment.

@@

A power analysis using G\*Power indicated that a sample size of `r ceiling((Exp2APA$v)+(Exp2APA$u)+1)` would allow us to detect a medium effect size ($f^2$ = `r Exp2APA$f2`), with three predictors (trait attribution; trait desirability; attribution*desirability), an alpha of `r Exp2APA$sig.level` and power of `r Exp2APA$power`. To counterbalance the sample equally and allow for the exclusion of outliers, a sample of `r length(Study2A_numeric$CaseID)` was collected. Within the sample, `r genderTable2A$Freq[1]` identified as female, `r genderTable2A$Freq[2]` as male, and `r genderTable2A$Freq[3]` as non-binary. Ages ranged from `r min(Study2A_numeric$Age)` to `r max(Study2A_numeric$Age)` ($M$ = `r round (mean(Study2A_numeric$Age), digits = 2)`, $SD$ = `r round (sd(Study2A_numeric$Age), digits = 2)`). Similar to Studies 1A and 1B, two attention checks were included in this study. As per our preregistered plan, 16 participants failed one or both attention checks; thus, their data were omitted and replaced. Participation took an average of nearly eight minutes.

@@

This study utilised a mixed design. All participants responded to items designed to measure animalistic dehumanization and mechanistic dehumanization. The desirability of the traits rated by participants was manipulated between subjects: half of the participants rated criminals on the extent to which they possessed socially desirable traits, and half rated criminals on the extent to which they possessed undesirable traits. All participants responded to a single item measuring the harshness of punishment endorsed.

@@

**Trait attribution**. After reading each vignette, participants responded to an 8-item scale measuring animalistic dehumanization (4 items) and mechanistic dehumanization (4 items) of the criminal portrayed. Participants made trait attributions by indicating their agreement with each item using an unmarked sliding scale ranging from 0 ('*Strongly Disagree*') to 100 ('*Strongly Agree*'), all of which were initially positioned at the scale's midpoint. Depending on the condition, the eight trait items were either socially desirable (*uniquely human: cultured, civilised, sophisticated, moral; human nature: generous, open-minded, warm, kind*) or socially undesirable (*uniquely human: corrupt, controlling, arrogant, bitter; human nature: jealous selfish, spiteful, cruel*). The lower the score, the more participants dehumanize the criminal target by denying them human traits.

@@

# plot(UH_harm, 5) #looks like there some outliers (over 3 standardised residuals)
#And, several scores have high leverage 2(3+1)/130 = .06.

@@

# dev.off()
# plot(HN_harm, 5) #looks like there some outliers (over 3 standardised residuals)
#And, several scores have high leverage 2(3+1)/130 = .06.

@@

This study utilised a mixed-methods design, matching that of Study 2A. The same attention checks used in Studies 1A, 1B and 2A were used in Study 2B.

@@

In Study 3A, we used an experimental design to examine further the hypothesised causal relationship between trait-based dehumanization and punishment when controlling for the social desirability of human traits incorporated into the stimuli. We described criminals with traits that varied in desirability and perceived humanness creating a $2 \times 2$ design. We then measured participants' willingness to endorse parole for each criminal described. The dual model predicts that criminals who are described in uniquely human terms will be more likely to be granted parole. We predicted that criminals described in socially desirable terms will be more likely to be granted parole. In principle, this design allows us to detect independent effects of dehumanization and trait sociability or an interaction between the two. In Study 3A, we specifically measure the extent to which animalistic dehumanization is causally related to parole decisions. Thus, we included uniquely human traits and those shared with other animals in our measures. We predict that participants will be more likely to endorse parole for criminals described with socially desirable traits, regardless of whether or not those traits are uniquely human or shared with other animals.

@@

**Vignettes**. All participants responded to the same four vignettes, each detailing a different scenario in which a criminal's eligibility for parole was assessed. Efforts were made to ensure that all four vignettes were similar in length, degree of detail, and contextual aspects, such as how long the criminal had spent in prison and who was described as attributing the traits to the criminal. In each vignette, the criminal's name and pronouns were gender-neutral, their age and ethnicity were not indicated, and their crime and sentence were not specified. The four vignettes can be seen in the Supplementary materials. In the uniquely human socially desirable condition, the criminal was described as *cultured, civilised, sophisticated*, and *moral*, while in the uniquely human condition socially undesirable, the criminal was described as *corrupt, controlling, arrogant* and *bitter*. In the animalistic desirable condition, the criminal was described as *energetic, trusting, genuine* and *having curiosity*, while in the animalistic undesirable condition, the criminal was described as *uncultured, unrefined, unsophisticated*, and *stupid*.

>>>>

@Bastian2013 conducted influential empirical studies testing the hypothesized association between the denial of human character traits and the endorsement of harsh punishment [see also @Barber2022; @Chen-Xia2023; @Kasper2022; @Morehouse2023; @Rousseau2023; @West2022]. The researchers measured how trait-based dehumanization influenced participants' punishment of criminals. Participants were asked to rate their agreement with four items assessing animalistic dehumanization of criminals: "*I felt like the person in the story was refined and cultured*" [reversed], "*I felt like the person in the story was rational and logical, like they were intelligent*" [reversed] "*I felt like the person in the story lacked self-restraint, like an animal*", and "*I felt like the person in the story was unsophisticated*". Participants were also asked to rate their agreement with four items assessing mechanistic dehumanization of criminals: "*I felt like the person in the story was open minded, like they could think clearly about things*" [reversed], "*I felt like the person in the story was emotional, like they were responsive and warm*" [reversed], "*I felt like the person in the story was superficial like they had no depth*", "*I felt like the person in the story was mechanical and cold, like a robot*". Bastian and colleagues reported that both forms of dehumanization predicted endorsement of harsh punishment for the criminals portrayed in their stimuli, concluding that their participants viewed criminals as '*subhuman and beastly*' (pp.9).

@@

@Bastian2013 seek to account for this possibility by statistically controlling for participants’ moral outrage at the targets’ behaviour in their analysis.  They report that the relationship between trait-based dehumanization and punishment remains even when moral outrage is controlled for.  While this is interesting and suggestive of the independent effects of dehumanization, it cannot fully address the conceptual weaknesses in how dehumanization was operationalized.  A more convincing way to de-confound evidence for trait-based dehumanization from evidence of negative evaluation is to ask participants to rate the target group on traits that are uniquely human but vary from socially desirable to undesirable [@Over2021a, @Over2021b].  Previous research conducted by @Enock2021b has established that undesirable character traits such as jealous, spiteful and bitter are considered unique to humans and socially undesirable.  Across three intergroup contexts, the researchers found that participants attributed socially desirable human traits more strongly to the ingroup and socially undesirable traits more strongly to the outgroup [see also @Decker2023; @Enock2021a; @Enock2022; @Enock2023].  @Enock2021b concluded that intergroup preference may better explain apparent evidence for trait-based dehumanization.  However, it is not yet clear how the attribution of uniquely human character traits relates to harm.  Addressing this question is crucial to understanding the extent to which the dual model of dehumanization can help explain real-world discrimination and negativity.

We revisit the hypothesized causal relationship between trait-based dehumanization and harm in the context of endorsing harsh punishment for criminals. In Studies 1A and 1B, we seek to conceptually replicate the key findings of Bastian and colleagues [-@Bastian2013], suggesting that the extent to which participants animalistically (Study 1A) and mechanistically (Study 1B) dehumanize criminals predicts the severity of the punishment participants endorse for them. In Studies 2A and 2B, we adopt a similar design but incorporate socially undesirable traits into our stimulus set. This addition to the design allows us to investigate whether trait-based dehumanization, undesirable trait attribution, or both predict the severity of punishment. Following Bastian and colleagues, and to understand the generalisability of our findings, we investigate these questions in relation to two different types of crime (violent crime and theft). In Studies 3A and 3B, we seek to investigate a similar question using an experimental design. We present participants with vignettes in which criminals are described using character traits that differ in how socially desirable they are and whether or not they are unique to humans. We also measure how these varying descriptions influence participants' parole decisions. This allows us to directly measure whether there is a causal relationship between trait-based dehumanization and punishment, independent of ingroup preference.

@@

@Bastian2013 presented evidence that the more participants dehumanize violent criminals, the harsher the punishment participants endorse for them. We sought to test whether we could conceptually replicate this relationship between trait-based dehumanization and punishment using terms very similar to those used by Bastian and colleagues. In study 1A, participants read a series of scenarios describing fictitious criminals and their violent crimes. Following this, participants rated the extent to which the criminals possessed four character traits that distinguish humans from animals (*refined, rational and logical, has a sense of morality* and *civilized*), which we refer to as uniquely human traits. Participants also rated the extent to which criminals possessed four character traits that distinguish humans from machines (*open-minded*, *emotionally responsive*, *has a depth of character*, and *interpersonally warm*), which we refer to as human nature traits. In investigating traits that distinguish humans from animals and machines, we are respecting the distinction between animalistic dehumanization and mechanistic dehumanization which is basic to the dual model [@Haslam2006, pp.256]. Participants also responded to an item measuring how harsh they thought the violent criminals' punishment should be. Following @Bastian2013, we predicted that the less participants attributed these uniquely human traits to criminals, the harsher the punishment they would recommend for criminals.

@@

**Trait attribution**. After reading each vignette, participants responded to items designed to measure trait-based dehumanization, broadly following the procedure of Bastian and colleagues [-@Bastian2013]. Participants indicated the extent to which they attributed four uniquely human traits (*refined, rational and logical, has a sense of morality* and *civilized*) and four human nature traits (*open-minded, emotionally responsive, has a depth of character*, and *interpersonally warm*) to the criminals depicted. Participants indicated their agreement with each item (e.g., '*I think* [e.g., *Charlie*] *is refined*') using an unmarked sliding scale from 0 ('*Strongly Disagree*') to 100 ('*Strongly Agree*'), with the sliders initially fixed at the midpoint. According to the dual model, lower scores indicated greater dehumanization of violent criminals. An attention check appeared halfway through the dehumanization items for two criminals ('*Please move the slider all the way to Strongly Agree/Disagree*').

@@

Following @Bastian2013, we utilized a within-subjects, correlational design. All participants read the same five vignettes presented in a random order and responded to the same trait attribution and punishment items. Participants' scores for the trait attribution items and the endorsed harshness of punishment item were then averaged across scenarios. The presentation of the items in the trait attribution task was also randomized.

@@

In line with our pre-registered criteria, seven highly influential were omitted from the analysis (remaining sample *N* = 93). After calculating the average attribution score for human nature traits and punishment for each participant, we conducted a simple linear regression to test attribution of human nature traits predicted the harshness of punishment endorsed for violent criminals. A significant negative relationship was found, *b* = `r round(Exp1a2$coefficients[2], digits = 2)` [`r round(Exp1a2CIdf$X2.5[2], digits = 2)`, `r round(Exp1a2CIdf$X97.5[2], digits = 2)`], *t* = `r round(Exp1a2$coefficients[6], digits = 2)`, *p* `r format.pval(Exp1a2$coefficients[8], digits = 2, eps = 0.001, nsmall = 3)`; see Figure \@ref(fig:figure1). This relationship shows that greater mechanistic dehumanization (operationalized as the denial of human nature traits) was associated with the endorsement of harsher punishment. The model explains `r round(Exp1a2$r.squared, digits = 2)*100`% of the variance in the harshness of punishment scores, $R^2$ = `r round(Exp1a2$r.squared, digits = 2)`, *F*(`r round(Exp1a2$fstatistic[2], digits = 2)`, `r round(Exp1a2$fstatistic[3], digits = 2)`) = `r round(Exp1a2$fstatistic[1], digits = 2)`.

@@

Study 1B investigates whether the relationship found in Study 1A replicates when participants are asked to judge a different type of criminal activity. In Study 1B, we examined whether animalistic and mechanistic forms of dehumanization, as operationalized by @Bastian2013, are associated with the harshness of punishment endorsed for individuals who commit theft. The design, materials and analysis plan were similar to that used in Study 1A, except that the scenarios involved theft rather than violent crime. In investigating a different type of crime, we follow the example set by Bastian et al. and seek to understand the generalisability of our results.

@@

**Vignettes** All participants responded to the same five vignettes, each detailing a crime involving theft (see electronic supplementary materials). As in Study 1A, an effort was made to ensure the vignettes were similar in structure and amount of detail. Once again, all of the perpetrators had gender-neutral names. An example of one of the theft vignettes is as follows: "*Until their recent arrest, Charlie had worked as a till operator at a local charity shop supporting individuals experiencing homelessness. Charlie had been stealing cash amounts varying from £5 to £50 from the tills almost daily over a five-year period. Police revealed that Charlie had stolen several thousand pounds from the charity shop while working there*."

@@

#looks like there some outliers (over 3 standardized residuals)
#several scores have high leverage 2(3+1)/130 = .06.

@@

#plot(HN_harm, 5) #looks like there some outliers (over 3 standardized residuals)
#several scores have high leverage 2(3+1)/130 = .06.

@@

#Rename The columns in both data frames so they match and allow for merging
names(Study2A_processed) <-c("CaseID", "Desirability_condition", "Consent", "English_Fluency", "Over_18", 
                             "Age", "Gender", "Gender_text", "Been_in_prison", "Criminal1_UH1_cultured.corrupt", 
                             "Criminal1_UH2_civilized.controlling", "Criminal1_UH3_sophisticated.arrogant", 
                             "Criminal1_UH4_moral.bitter", "Criminal1_HN1_generous.jealous", "Criminal1_HN2_open-minded.selfish", 
                             "Criminal1_HN3_warm.spiteful", "Criminal1_HN4_kind.cruel", "Criminal1_harm", 
                             "Criminal2_UH1_cultured.corrupt", "Criminal2_UH2_civilized.controlling", 
                             "Criminal2_UH3_sophisticated.arrogant", "Criminal2_UH4_moral.bitter", 
                             "Criminal2_HN1_generous.jealous", "Criminal2_HN2_open-minded.selfish", "Criminal2_HN3_warm.spiteful", 
                             "Criminal2_HN4_kind.cruel", "Criminal2_harm", "Criminal3_UH1_cultured.corrupt", 
                             "Criminal3_UH2_civilized.controlling", "Criminal3_UH3_sophisticated.arrogant", 
                             "Criminal3_UH4_moral.bitter", "Criminal3_HN1_generous.jealous", "Criminal3_HN2_open-minded.selfish", 
                             "Criminal3_HN3_warm.spiteful", "Criminal3_HN4_kind.cruel", "Criminal3_harm", "Criminal4_UH1_cultured.corrupt", 
                             "Criminal4_UH2_civilized.controlling", "Criminal4_UH3_sophisticated.arrogant", "Criminal4_UH4_moral.bitter", 
                             "Criminal4_HN1_generous.jealous", "Criminal4_HN2_open-minded.selfish", "Criminal4_HN3_warm.spiteful", 
                             "Criminal4_HN4_kind.cruel", "Criminal4_harm", "Criminal5_UH1_cultured.corrupt", 
                             "Criminal5_UH2_civilized.controlling", "Criminal5_UH3_sophisticated.arrogant", 
                             "Criminal5_UH4_moral.bitter", "Criminal5_HN1_generous.jealous", "Criminal5_HN2_open-minded.selfish", 
                             "Criminal5_HN3_warm.spiteful", "Criminal5_HN4_kind.cruel", "Criminal5_harm", "UH_mean", "HN_mean", 
                             "Harm_mean", "Outlier_Animalistic_model", "Outlier_Mechanistic_model", "Study")

names(Study2B_processed) <-c("CaseID", "Desirability_condition", "Consent", "English_Fluency", "Over_18", 
                              "Age", "Gender", "Gender_text", "Been_in_prison", "Criminal1_UH1_cultured.corrupt", 
                              "Criminal1_UH2_civilized.controlling", "Criminal1_UH3_sophisticated.arrogant", 
                              "Criminal1_UH4_moral.bitter", "Criminal1_HN1_generous.jealous", "Criminal1_HN2_open-minded.selfish", 
                              "Criminal1_HN3_warm.spiteful", "Criminal1_HN4_kind.cruel", "Criminal1_harm", 
                              "Criminal2_UH1_cultured.corrupt", "Criminal2_UH2_civilized.controlling", 
                              "Criminal2_UH3_sophisticated.arrogant", "Criminal2_UH4_moral.bitter", 
                              "Criminal2_HN1_generous.jealous", "Criminal2_HN2_open-minded.selfish", "Criminal2_HN3_warm.spiteful", 
                              "Criminal2_HN4_kind.cruel", "Criminal2_harm", "Criminal3_UH1_cultured.corrupt", 
                              "Criminal3_UH2_civilized.controlling", "Criminal3_UH3_sophisticated.arrogant", 
                              "Criminal3_UH4_moral.bitter", "Criminal3_HN1_generous.jealous", "Criminal3_HN2_open-minded.selfish", 
                              "Criminal3_HN3_warm.spiteful", "Criminal3_HN4_kind.cruel", "Criminal3_harm", "Criminal4_UH1_cultured.corrupt", 
                              "Criminal4_UH2_civilized.controlling", "Criminal4_UH3_sophisticated.arrogant", "Criminal4_UH4_moral.bitter", 
                              "Criminal4_HN1_generous.jealous", "Criminal4_HN2_open-minded.selfish", "Criminal4_HN3_warm.spiteful", 
                              "Criminal4_HN4_kind.cruel", "Criminal4_harm", "Criminal5_UH1_cultured.corrupt", 
                              "Criminal5_UH2_civilized.controlling", "Criminal5_UH3_sophisticated.arrogant", 
                              "Criminal5_UH4_moral.bitter", "Criminal5_HN1_generous.jealous", "Criminal5_HN2_open-minded.selfish", 
                              "Criminal5_HN3_warm.spiteful", "Criminal5_HN4_kind.cruel", "Criminal5_harm", "UH_mean", "HN_mean", 
                              "Harm_mean", "Outlier_Animalistic_model", "Outlier_Mechanistic_model", "Study")

@@

Study 2A investigated whether apparent evidence for a relationship between trait-based dehumanization and endorsement of harsh punishment for violent criminals remains when controlling for the desirability of the traits. We tested this by introducing character traits perceived as uniquely human yet socially undesirable into the stimulus set [@Enock2021b; @Over2021a; @Over2021b]. The dual model predicts that to the extent criminals are denied uniquely human character traits, they will be subjected to harsher punishment. We predict that trait desirability will moderate the relationship between human trait attribution and punishment. More specifically, we predict that the extent to which violent criminals are denied socially desirable character traits, and attributed socially undesirable character traits, will predict harsh punishment.

@@

A power analysis using G\*Power indicated that a sample size of `r ceiling((Exp2APA$v)+(Exp2APA$u)+1)` would allow us to detect a medium effect size ($f^2$ = `r Exp2APA$f2`), with three predictors (trait attribution; trait desirability; attribution*desirability), an alpha of `r Exp2APA$sig.level` and power of `r Exp2APA$power`. To counterbalance the sample equally and allow for the exclusion of outliers, a sample of `r length(Study2A_numeric$CaseID)` was collected. Within the sample, `r genderTable2A$Freq[1]` identified as female, `r genderTable2A$Freq[2]` as male, and `r genderTable2A$Freq[3]` as non-binary. Ages ranged from `r min(Study2A_numeric$Age)` to `r max(Study2A_numeric$Age)` ($M$ = `r round (mean(Study2A_numeric$Age), digits = 2)`, $SD$ = `r round (sd(Study2A_numeric$Age), digits = 2)`). Similar to Studies 1A and 1B, two attention checks were included in this study. As per our pre-registered plan, 16 participants failed one or both attention checks; thus, their data were omitted and replaced. Participation took an average of nearly eight minutes.

@@

This study utilized a mixed design. All participants responded to items designed to measure animalistic dehumanization and mechanistic dehumanization. The desirability of the traits rated by participants was manipulated between subjects: half of the participants rated criminals on the extent to which they possessed socially desirable traits, and half rated criminals on the extent to which they possessed undesirable traits. All participants responded to a single item measuring the harshness of punishment endorsed.

@@

**Trait attribution**. After reading each vignette, participants responded to an 8-item scale measuring animalistic dehumanization (4 items) and mechanistic dehumanization (4 items) of the criminal portrayed. Participants made trait attributions by indicating their agreement with each item using an unmarked sliding scale ranging from 0 ('*Strongly Disagree*') to 100 ('*Strongly Agree*'), all of which were initially positioned at the scale's midpoint. Depending on the condition, the eight trait items were either socially desirable (*uniquely human: cultured, civilized, sophisticated, moral; human nature: generous, open-minded, warm, kind*) or socially undesirable (*uniquely human: corrupt, controlling, arrogant, bitter; human nature: jealous selfish, spiteful, cruel*). The lower the score, the more participants dehumanize the criminal target by denying them human traits.

@@

# plot(UH_harm, 5) #looks like there some outliers (over 3 standardized residuals)
#And, several scores have high leverage 2(3+1)/130 = .06.

@@

# dev.off()
# plot(HN_harm, 5) #looks like there some outliers (over 3 standardized residuals)
#And, several scores have high leverage 2(3+1)/130 = .06.

@@

This study utilized a mixed-methods design, matching that of Study 2A. The same attention checks used in Studies 1A, 1B and 2A were used in Study 2B.

@@

In Study 3A, we used an experimental design to examine further the hypothesized causal relationship between trait-based dehumanization and punishment when controlling for the social desirability of human traits incorporated into the stimuli. We described criminals with traits that varied in desirability and perceived humanness creating a $2 \times 2$ design. We then measured participants' willingness to endorse parole for each criminal described. The dual model predicts that criminals who are described in uniquely human terms will be more likely to be granted parole. We predicted that criminals described in socially desirable terms will be more likely to be granted parole. In principle, this design allows us to detect independent effects of dehumanization and trait sociability or an interaction between the two. In Study 3A, we specifically measure the extent to which animalistic dehumanization is causally related to parole decisions. Thus, we included uniquely human traits and those shared with other animals in our measures. We predict that participants will be more likely to endorse parole for criminals described with socially desirable traits, regardless of whether or not those traits are uniquely human or shared with other animals.

@@

**Vignettes**. All participants responded to the same four vignettes, each detailing a different scenario in which a criminal's eligibility for parole was assessed. Efforts were made to ensure that all four vignettes were similar in length, degree of detail, and contextual aspects, such as how long the criminal had spent in prison and who was described as attributing the traits to the criminal. In each vignette, the criminal's name and pronouns were gender-neutral, their age and ethnicity were not indicated, and their crime and sentence were not specified. The four vignettes can be seen in the Supplementary materials. In the uniquely human socially desirable condition, the criminal was described as *cultured, civilized, sophisticated*, and *moral*, while in the uniquely human condition socially undesirable, the criminal was described as *corrupt, controlling, arrogant* and *bitter*. In the animalistic desirable condition, the criminal was described as *energetic, trusting, genuine* and *having curiosity*, while in the animalistic undesirable condition, the criminal was described as *uncultured, unrefined, unsophisticated*, and *stupid*.
